\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{parskip}
\usepackage{float}

\begin{document}
\raggedbottom
\title{ArtBench-10 Generative Modeling: A Comparative Study of VAE, GAN, and Diffusion Models}
\author{Tiago Silva \and João Coelho}
\institute{University of Coimbra}
\maketitle

\begin{abstract}
    Nothing yet.
\end{abstract}

\section{Introduction}

\section{Problem Formulation}


\section{Methodological Framework}

\subsection{Implementation Details and Reproducibility}

To ensure reproducibility, we adopted a deterministic seeding policy across the pipeline. 
In the base training workflow, all stochastic operations were initialized with seed 42. 
During hyperparameter optimization, each Optuna trial used a deterministic trial-specific 
seed ($42 + \text{trial\_id}$), and the train/validation split was generated with a fixed seed. 
Final evaluation was conducted across 10 distinct seeds, followed by reporting mean and standard deviation
of metrics. During the experimentation phase, a 20\%  subset (10,000 images) was used for hyperparameter sweeps via Bayesian optimization. 
Once optimal parameters werei  dentified, the final models were retrained on the 
full 50,000-image training set.

Regarding data preprocessing, all images were resized and center-cropped to $32\times32$, 
converted to tensors, and normalized channel-wise with mean $(0.5, 0.5, 0.5)$ and 
standard deviation $(0.5, 0.5, 0.5)$, resulting in inputs scaled to $[-1,1]$. 
This normalization is important in our setting because it aligns the input scale with the 
output range typically used by generative architectures (e.g, tanh outputs values in this range), improves optimization conditioning, 
and contributes to more stable gradient dynamics, particularly for adversarial and 
diffusion-based training (forward noising process is defined around 0 mean Gaussian noise).

All training functions followed the same execution template:
\begin{itemize}
    \item Initialization of models, output directories, and training variables;
    \item Main training loop for 50 epochs, with checkpoints and generated samples saved every 10 epochs;
    \item Saving of the final model.
\end{itemize}

\subsubsection{Variational Autoencoder}
For this model, we implemented a symmetric 4 layer convolutional design 
(32, 16, 8, 4, 2) with leak Relu activations and batch normalization 
after each convolutional layer. Following ~\cite{vae} we also
implemented the reparameterization trick to allow backpropagation through the stochastic sampling of the latent space.

\subsubsection{Generative Adversarial Network}
For the adversarial baseline, we implemented a Deep Convolutional GAN (DCGAN) with a 
transposed-convolution generator and a strided-convolution discriminator just as dicussed in
the PL classes with some notations coming from~\cite{dcgan}. 
The generator maps a latent vector $z$ to a $32\times32\times3$ image through 
four upsampling blocks (ConvTranspose2d + BatchNorm + ReLU), followed by a \texttt{tanh} output layer. 
The discriminator mirrors this design with four convolutional blocks using LeakyReLU activations and 
BatchNorm in intermediate layers, ending in a sigmoid probability for binary real/fake classification.

Training follows the standard alternating minimax procedure with binary cross-entropy loss: 
the discriminator is first updated using real images and generated images, and then the 
generator is updated to maximize the discriminator's probability of classifying generated 
samples as real. We used Adam optimization (with tuned learning rate and $\beta_1$, and 
fixed $\beta_2=0.999$), as well as one-sided label smoothing for real targets to reduce 
discriminator overconfidence and improve early training stability. 

\subsubsection{Latent Denoise Network}

The Latent Denoise Network model's architecture was the same as the one in the PL classes.
For the diffusion process, the DDPM scheduler used was the GaussianDiffusion model.
For the time steps embeddings, a Sinusoidal Position Embedding, with another 3 layers of Linear, SiLu and Linear transformations, respectivelly, were used.
For the reverse diffusion step, ResnetBlocks in sequence were used.
For the encoding and decoding of samples to and from the latent space, the VAE previously trained was used.

The only changes made to the provided code were to the number of latent channels, which had to be the same as the trained VAE latent channels', 
and the implementation of the sample, train step and image generation functions inside the class of the model.
These sample and train step functions were implemented just like the respective PL class demanded.

\subsubsection{Pixel UNet}

The Pixel UNet model's architecture, just like the previous model, was the same as the one in the Diffusion models's PL class.
The diffusion process and the time steps embeddings were the same as the ones in the previous model.
The only differences were in the reverse diffusion step, where 6 ResnetBlocks were used (2 in downscaling, 2 keeping the resolution and 2 in upscaling), 
and the lack of a latent space.

Similarly to the Latent Denoise Network, the main changes to the code were with the implementation of the sample, train step and image generation inside the class of the model.
These functions are very similar to the ones implemented with the previous model, only with slight adaptations.
The only other difference from the PL class's code was the change in the input channels, which became 3, due to the dataset images being represented in the RGB space.

\subsection{Theoretical Rationale for Hyperparameter Choices}
The choice of which hyperparameters to tune and the search space design follows 4 main principles:
enable sufficient representational capacity to model ArtBench-10's complexity, stable training dynamics,
regularization to prevent overfitting, and computational feasibility under resource constraints. 

\subsubsection{Variational Autoencoders (VAE)}
For the VAE we tuned latent dimensionality, base channels, learning rate, batch size, and $\beta$ (the weight of the KL divergence term in the loss).
Latent dimension is intended to control the capacity of capturing the underlying data distribution, expressing a trade-off between reconstruction fidelity and generalization.
Base channels determine the width of the convolutional layers, which affects the model's ability to learn complex features.
Learning rate was sampled from a log-uniform distribution to allow exploration across several orders of magnitude, which is crucial for stable convergence.
The $\beta$ parameter was also sampled from a log-uniform distribution to explore a wide range of regularization strengths, consistent with the objective function:
\begin{equation}
    \mathcal{L} = \mathbb{E}_{q(z|x)}[\log p(x|z)] - \beta \cdot D_{KL}(q(z|x) || p(z))
\end{equation}
where $\beta$ controls the trade-off between reconstruction accuracy and latent space regularization.

\subsubsection{Generative Adversarial Networks (GAN)}
For GANs, the search space targeted parameters known to affect training 
stability and sample diversity. The latent dimensionality and the number of 
feature maps determine generator and discriminator capacity. We additionally 
tuned the real-label target used in label smoothing, which can reduce 
discriminator overconfidence and improve gradient quality early in training. 
The Adam $\beta_1$ parameter was also tuned because adversarial optimization 
is sensitive to momentum settings.

\subsubsection{Diffusion Models}
For diffusion models, we tuned the denoiser width and depth 
(\texttt{model\_channels} and \texttt{num\_res\_blocks}), learning rate, 
batch size, and the number of diffusion timesteps. The timestep count 
captures the trade-off between computational cost and generative fidelity: 
more timesteps generally improve sample quality but increase training and 
sampling cost. In the latent-diffusion setting, the VAE was kept frozen so 
that the study isolates the effect of denoiser hyperparameters and avoids 
conflating them with representation learning in the autoencoder.



\section{Experimental Design}
\subsection{Hyperparameter Optimization Strategy}
We performed hyperparameter optimization with Bayesian optimization (Optuna). To keep the search feasible on standard hardware, 
tuning was conducted on a development subset comprising 20\% of the training data. For each model family, candidate configurations 
were trained on a training split and selected using performance on a held-out validation split. We minimized the following 
surrogate objectives: (i) \textit{VAE}: validation reconstruction loss; (ii) \textit{GAN}-based models: validation sum of discriminator 
and generator losses; (iii) \textit{Diffusion models}: validation denoising loss under a fixed Gaussian noise schedule.
The hyperparameter search spaces are summarized in Table~\ref{tab:hpo_spaces}, and the best configurations found on the studies are 
reported in Table~\ref{tab:hpo_best}. We used validation losses as surrogates during tuning because FID/KID are computationally expensive 
and high-variance at the per-trial scale: they require generating large sample sets and repeated feature extraction. In contrast, validation losses are 
available at each epoch and exhibit lower variance which improves sample efficiency under limited computational resources. We note, however, that GAN losses are far from beeing perfect proxies 
for perceptual sample quality; therefore, we use this objective only for hyperparameter selection during tuning and rely on qualitative inspection 
for the selected final parameters.
\begin{table}[H]
    \centering
    \caption{Hyperparameter search spaces used in Bayesian optimization.}
    \label{tab:hpo_spaces}
    \begin{tabular}{p{0.2\textwidth}p{0.3\textwidth}p{0.5\textwidth}}
        \toprule
        \textbf{Model} & \textbf{Parameters} & \textbf{Values / Range} \\
        \midrule
        VAE &
        latent\_dim, base\_channels, batch\_size &
        $\{16,32,64\}$; $\{16,32,64\}$; $\{32,64,128\}$ \\
        & lr, $\beta$ &
        $[10^{-4}, 3\times 10^{-3}]$ log-uniform; $[10^{-6}, 10^{-3}]$ log-uniform \\
        \midrule
        DCGAN &
        latent\_dim, feature\_maps, batch\_size &
        $\{128,256,512\}$; $\{32,64\}$; $\{32,64,128\}$ \\
        & lr, $\beta_1$, real\_label &
        $[10^{-4}, 5\times 10^{-4}]$ log-uniform; $[0.3,0.7]$; $[0.85,1.0]$ \\
        \midrule
        PixelUNet &
        model\_channels, batch\_size, num\_timesteps &
        $\{32,64,96\}$; $\{16,32,64\}$; $\{500,1000\}$ \\
        & lr &
        $[10^{-4}, 5\times 10^{-4}]$ log-uniform \\
        \midrule
        LatentDenoiser &
        model\_channels, num\_res\_blocks, batch\_size &
        $\{32,64,96\}$; $\{2,3,4\}$; $\{16,32,64\}$ \\
        & lr, num\_timesteps &
        $[10^{-4}, 5\times 10^{-4}]$ log-uniform; $\{500,1000\}$ \\
        \bottomrule
    \end{tabular}
\end{table}



\subsection{Evaluation Protocol}
Across the selected configurations (Table~\ref{tab:hpo_best}), 
the optimization process consistently favored smaller batch sizes. 
This aligns with the well-documented generalization gap observed in 
large-batch training~\cite{keskar}. Specifically, the inherent noise 
in small-batch stochastic gradients serves as a form of implicit 
regularization, facilitating the escape from sharp local minima and 
leading the model toward flatter, more robust regions of the loss 
landscape. Regarding the learning rates, the observed values
are relatively low, consistent with the need for stable convergence and
stability during training, especially for GANs which are known to be sensitive to hyperparameter settings.
Latent dimensions reveal important differences across model families: the VAE favored a much smaller 
latent space (32) compared to the DCGAN (256), which may reflect the different in the 
model's architectures: GANs don't pay an explicity penalty (KL divergence)
for using larger latent spaces, while VAEs are encouraged to find compact representations.
A larger latent space may encourage more diverse samples as long as the generator
actually uses it. For the VAE, the optimal $\beta$ value is very low, 
indicating that the model prioritizes reconstruction fidelity over latent space regularization.
For the DCGAN, the optimal $\beta_1$ value is around 0.48, which is close 
to the commonly used value of 0.5~\cite{dcgan}, suggesting a balance between stability and convergence speed. 
The real\_label parameter being close to 1 (0.9836) indicates that the discriminator 
benefits from a small anti-overconfidence margin, which can help prevent it from becoming too confident 
and overpowering the generator during training without weakening the learning signal too much.


\begin{table}[H]
    \centering
    \caption{Best configurations obtained.}
    \label{tab:hpo_best}
    \begin{tabular}{p{0.28\textwidth}cccc}
        \toprule
        \textbf{Parameter} & \textbf{VAE} & \textbf{DCGAN} & \textbf{PixelUnet} & \textbf{LatentDenoiser}\\
        \midrule
        Best objective value & 0.02658 & 1.68158 & 0.0358 & 0.2598 \\
        Learning rate & 0.0005317 & 0.0001261 & 0.0004748 & 0.0001769\\
        Batch size & 32 & 32 & 16 & 32 \\
        Latent dimension & 32 & 256 & 64 & 64 \\
        Base channels / feature maps & 64 & 64 & 64 & 64 \\
        $\beta$ (VAE) & $1.14 \times 10^{-6}$ & -- & -- & -- \\
        $\beta_1$ (Adam, DCGAN) & -- & 0.4813 & -- & -- \\
        real\_label (DCGAN) & -- & 0.9836 & -- & -- \\
        num\_timesteps (diffusion) & -- & -- & 1000 & 1000 \\
        num\_res\_blocks (LatentDenoiser) & -- & -- & -- & 4 \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Results and Interpretation}

\subsection{Quantitative Results}
Table~\ref{tab:summary_results} summarizes the quantitative evaluation across model families using 5{,}000 generated 
samples and 10 random seeds per model. We report mean $\pm$ standard deviation across seeds. 
Arrows indicate whether lower ($\downarrow$) or higher ($\uparrow$) values are preferred. 
The best value in each metric is highlighted in bold.

\begin{table}[H]
    \centering
    \caption{Summary of quantitative results (mean $\pm$ std across seeds).}
    \label{tab:summary_results}
    \begin{tabular}{lcccc}
        \toprule
        \textbf{Model} & \textbf{FID $\downarrow$} & \textbf{KID $\downarrow$} & \textbf{Precision $\uparrow$} & \\
        \midrule
        VAE & 134.228 $\pm$ 0.531 & 0.13136 $\pm$ 0.00109 & 0.80548 $\pm$ 0.00671\\
        DCGAN & \textbf{101.248 $\pm$ 0.864} & \textbf{0.07664 $\pm$ 0.00153} & 0.86400 $\pm$ 0.00489 \\
        PixelUNet & 116.933 $\pm$ 0.708 & 0.09972 $\pm$ 0.00107 & \textbf{0.90804 $\pm$ 0.00512} \\
        LatentDenoiser & 119.271 $\pm$ 0.416 & 0.10068 $\pm$ 0.00122 & 0.78472 $\pm$ 0.00466 \\
        \bottomrule
    \end{tabular}
\end{table}

Overall, DCGAN achieves the best performence in terms of 
global distributional alignment obtaining the lowest FID (101.248 $\pm$ 0.864)
and KID (0.07664 $\pm$ 0.00153) scores. This suggests that the DCGAN is more effective at 
capturing the overall data distribution compared to the other models. The VAE, on the 
other hand, has the worst performance in terms of FID and KID, 
which is consistent with the known limitations of VAEs in generating overly
smooth reconstructions due to the variational objective, which
can lead to blurry samples and less realistic outputs.
PixelUNet achieves the highest precision (0.90804 $\pm$ 0.00512), indicating 
that it generates a higher proportion of samples that are close to the 
real data manifold. However, the higher FID and KID values
compared to DCGAN suggest that the improvement in sample realism 
does not necessarily translate to better overall distributional alignment.
They, therefore, generate more realistic samples but with less diversity.
The LatentDenoiser achieves intermediate FID and KID values but the 
lowest precision among the evaluated models. This suggests that while 
its generated distribution is moderately aligned with the target dataset 
at a global level, a larger fraction of its individual samples fall outside 
the real data manifold compared to the other architectures. Such behavior may 
indicate incomplete convergence or reduced stability in the denoising process 
during sampling. In addition, the small standard deviations across seeds for all 
models suggest that the results are consistent and not heavily influenced by
random initialization or stochastic training dynamics which is a positive sign of the 
robustness of the training process.


\section{Conclusion}

\section*{Use of External Resources and AI Tools}
This project utilized ArtBench-10 (Kaggle). Gemini assisted in LaTeX formatting and analysis verification.

\nocite{artbench,vae,dcgan,fid,keskar}
\bibliographystyle{splncs04}
\bibliography{main}

\end{document}
